Repeatability of Pre-Flashover Fire Patterns on Gypsum Wall Board

### 2015-03-08 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Editing of references.

A subset of the intentional fires are considered arson.  Arson as defined by the U.S. Federal Bureau of Investigation (FBI) Uniform Crime Reporting (UCR) Program defines arson as any willful or malicious burning or attempting to burn, with or without intent to defraud, a dwelling house, public building, motor vehicle or aircraft, personal property of another, etc.~\cite{Crime}.

@@

Fires are investigated in order to determine the ``origin and cause'' of the fire and in cases of arson there is an effort to determine who was responsible for setting the fire.  This investigation process must follow the scientific method as documented in NFPA 921~\cite{NFPA_921}.  There are numerous textbooks on the subject of conducting a fire investigation~\cite{Almirall:2004,Fire_Investigation,DeHaan:2012,Icove:2013,Lentinin:2006,Noon:1995}. The NFPA guide and the texts provide information to a fire investigator collecting data from the scene, analyzing the data, and developing a hypothesis about the fire.

Patterns produced by the fire are in many cases a significant portion of the data collected and are analyzed at the scene to determine the area of origin.  One of the basic methods of documenting the fire scene is to photograph fire patterns.  As noted by Icove, DeHaan, and Haynes  ``'the ability to document and interpret fire patterns accurately is essential to investigators reconstructing fire scenes…''. [13]  A fire pattern is defined in NFPA 921 as, “the visible or measurable physical change, or identifiable shapes, formed by a fire effect or group of fire effects” [921].
DeHaan [12] categorizes the fire effects that form patterns as follows:
\begin{enumerate}
\item Surface deposits – no irreversible effect on surface
\item Surface thermal effects – physical change such a discoloration or melting
\item Charring –evidence of surface burning
\item Penetration – charring below the surface
\item Consumption –loss of surface material, charring throughout
\end{enumerate}
Lentini points out that most fire patterns are generated and later observed on two-dimensional surfaces at the place where those surfaces intersect with the three-dimensional fire. Various [14] types of fire patterns, such as; ``V-shaped'', ``hour-glass'', and ``inverted cone'', have come from common observation at actual fire scenes.  As a result, the observations are typically qualitative in nature.

Previous fire pattern research by the National Institute of Justice (NIJ), the National Institute of Standards and Technology (NIST), and the United States Fire Administration (USFA) has shown that fire patterns provide data useful for the determination of fire origin.  The reports noted the impact of ventilation on the development of the burn patterns. [16, 17] A large number of other factors affect the formation of these patterns: burn time, heat release rate of the fire source, fire exposure, target fuel composition, adjacent fuel(s) and compartmentation, to name a few.    Given the limited number of experiments in the literature and the large number of variables, it has been difficult to fully understand a cause and effect relationship between the fire scenarios and the resulting patterns.  Therefore, fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause.[18]  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs).  FDTs incorporates references to the source documents and experimental work upon which the correlations are based. In addition, NRC tasked NIST with conducting a verification and validation study of the correlations. Overholt developed the NIST verification and validation study.[19] In a private communication, Overholt noted that not all of the correlations in FDTs were included in the study since a sufficient amount of reliable data in the literature to not available to conduct validation.  Two of the types of correlations, which could not be validated, were flame height and time to ignition.[19]

A survey of zone and field fire models was compiled and is posted on www.firemodelsurvey.com.[20] The survey was updated in 2014, but it includes models that were documented in the first fire model survey conducted in 1992. As a result many of the models are listed for historical purposes only.  There are 53 zone models listed. Only 3 of the 53 are being actively supported or maintained.  The survey also includes 22 field models.  10 of the 22 field models are listed as being actively supported.  All of the correlations and models have been to some extent been validated or checked against experimental data for a limited set of scenarios. [21]

Of the active field models, FDS version 6 has one of the most rigorous and publicly accessible verification and validation programs.[22] In the current FDS validation guide more than 45 experimental data sets are used to quantify the prediction capabilities of the model.  Results are presented in terms of model bias and variation to inform users of whether particular quantities (temperature, heat flux, pressure, etc.) are over or under predicted compared to experimental data.

However for fire models to be demonstrated as reliable and suitable for use in court, studies such as the one conducted here need to be conducted to provide the scientific basis for the use of fire models for re-construction.  The research results may enable fire investigation teams to have a better understanding of the fire scene and enable improved support of their hypothesis on the area of origin based on fire patterns.  This idea was discussed when the Fire Protection Research Foundation convened a Research Advisory Council on Post-fire Analysis in 2002.  Recommendations for research and development in their white paper included, ``advance the capabilities of computational fluid dynamics fire modeling, particularly as applied to fire ignition scenarios and fire pattern development and interpretation.''  [23]

It has been pointed out in this introduction that fire is still a threat to life and property and how fire investigation plays a role in preventing future fires.  Given that fire pattern interpretation and analysis is a critical step in most fire investigations, it is surprising to find the limited amount or research and analysis tools directly related to this practice.  The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA 921, Guide for Fire and Explosion Investigations.  The first edition of NFPA 921 was issued in 1992.  As provided by the document scope, it `` is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.'' The document’s purpose includes the goal to be ``…a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.'' [8] However, NFPA 921 can only incorporate the fire science and research results that have been produced.  Although NFPA 921 is a ``guide'' as opposed to a ``standard'' is considered the standard of care for the fire investigation community in countries that recognize and use NFPA standards and guides.

As the result of the review of several arson cases, most notably the cases of Ernest Ray Willis and Cameron Todd Willingham, the practice of fire investigation as a forensic science was called into question.  Beyler conducted a review of both cases for the Texas Forensic Science Commission.  Both fires occurred prior to the release of NFPA 921 in 1992.  The Beyler analysis, issued in August of 2009, concluded that the investigations of both cases ``did not comport with either the modern standard of care expressed by NFPA 921, or the standard of care expressed by fire investigation texts and papers in the period 1980 – 1992.'' The fire investigators had a poor understanding of science and their findings of arson could not be sustained.[24]

Another landmark document regarding the state of forensic science in the United States was published in 2009.  The 328 page report, titled Strengthening Forensic Science in the United States: A Path Forward, by the U.S. National Academy of Sciences, reviewed 13 different forensic science disciplines including; biological evidence, friction ridge patterns, tool mark and firearms identification, and analysis of explosives evidence and fire debris, to name a few.  The review of the analysis of explosives evidence and fire debris is covered in three pages of the report.  The summary assessment of analysis of explosives evidence is positive, ``the scientific foundations exist to support the analysis of explosions, because such analysis is based primarily on well-established chemistry .''[25]  The complete summary assessment for fire debris analysis is included here: ``By contrast, much more research is needed on the natural variability of burn patterns and damage characteristics and how they are affected by the presence of various accelerants.  Despite the paucity of research, some arson investigators continue to make determinations about whether or not a particular fire was set.  However, according to testimony presented to the committee, many of the rules of thumb, that were typically assumed to indicate that an accelerant was used (e.g. ``alligatoring'' of wood, specific char patterns) have been shown not to be true.  Experiments should be designed to put arson investigations on a more solid scientific footing .'' [25]
This study is in response to the assessment above, specifically regarding the natural variability of fire patterns.

@@

Well-documented and characterized technologies such as thermocouples and heat flux gauges were used to assess the fire and the thermal exposure of the target material.  Oxygen consumption calorimetry and mass loss was used for the determining the heat release rate of the fires.

The most important measurements for the field investigator are the visual measurements that can be made at the scene or determined from scene photographs. Several methods of imaging were used for these experiments: still photography, digital video, and infrared video.  Computer based imaging techniques, such as those contained in the Image Stream program suite, were examined and adapted to analyze the flame height of the source fires [26].

@@

The fires were characterized in terms of heat release rate, mass loss rate, fire plume temperatures and emitted heat flux. Videos were recorded for each experiment.  This ``field'' data from the videos was  used to determine the flame heights. Experiments were conducted with three different fuel types: a gas fueled mass flow controlled burner, a liquid fueled burner, and a solid fuel.
Instrumented Calibration Target Wall – A free standing, non-combustible, calcium silicate wall was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.

@@

Comparative experiments were conducted to examine the impact of the method of construction and support of the walls and corner sections on the development of the burn patterns.  Three cases were examined: 1) 12.7~mm thick, regular core, painted gypsum board on the front of a wall frame (open back), 2) 12.7~mm thick, regular core, painted gypsum board on the front and back of a wall frame (interior wall), and 3) 12.7~mm thick, regular core, painted gypsum board on the front and back of a wall frame, with fiberglass batting insulation in the voids of the wood frame (exterior wall).  The main consideration was to identify any differences in the height, width, shape, and area of the fire patterns. The results of these experiments determined the construction of the walls and corners used in subsequent phases of this study.

@@

Several types of bench scale experiments were conducted to generate data. This data could be used to assist in the analysis of the fire patterns and potentially be used as input for the FDS models.
Critical Heat Flux – Conducted a series of experiments on samples of painted gypsum wallboard to determine the repeatability of a “critical heat flux” that results in a clear line of demarcation, a ``burn/no burn'' line.  The flooring radiant panel (ASTM E648) was used to examine this value.

@@

Cone calorimeter (ASTM 1354) experiments were conducted to estimate the ignition temperature, the time to ignition and the heat release rate from samples of painted gypsum wallboard under thermal flux conditions ranging from 25~kW/m$^2$ to 75~kW/m$^2$.

>>>>

A subset of the intentional fires are considered arson.  Arson as defined by the U.S. Federal Bureau of Investigation (FBI) Uniform Crime Reporting (UCR) Program defines arson as any willful or malicious burning or attempting to burn, with or without intent to defraud, a dwelling house, public building, motor vehicle or aircraft, personal property of another, etc.~\cite{Crime:2010}.

@@

Fires are investigated in order to determine the ``origin and cause'' of the fire and in cases of arson there is an effort to determine who was responsible for setting the fire.  This investigation process must follow the scientific method as documented in NFPA 921~\cite{NFPA_921}.  There are numerous textbooks on the subject of conducting a fire investigation~\cite{Almirall:2004,Fire_Investigation,DeHaan:2012,Icove:2013,Lentini:2006,Noon:1995}. The NFPA guide and the texts provide information to a fire investigator collecting data from the scene, analyzing the data, and developing a hypothesis about the fire.

Patterns produced by the fire are in many cases a significant portion of the data collected and are analyzed at the scene to determine the area of origin.  One of the basic methods of documenting the fire scene is to photograph fire patterns.  As noted by Icove, DeHaan, and Haynes  ``'the ability to document and interpret fire patterns accurately is essential to investigators reconstructing fire scenes…''~\cite{Icove:2013}.  A fire pattern is defined in NFPA~921 as, ``the visible or measurable physical change, or identifiable shapes, formed by a fire effect or group of fire effects''~\cite{NFPA_921}.

DeHaan~\cite{DeHaan:2012} categorizes the fire effects that form patterns as follows:
\begin{enumerate}
\item Surface deposits -– no irreversible effect on surface
\item Surface thermal effects –- physical change such a discoloration or melting
\item Charring -– evidence of surface burning
\item Penetration -– charring below the surface
\item Consumption –- loss of surface material, charring throughout
\end{enumerate}
Lentini points out that most fire patterns are generated and later observed on two-dimensional surfaces at the place where those surfaces intersect with the three-dimensional fire. Various types of fire patterns, such as; ``V-shaped'', ``hour-glass'', and ``inverted cone'', have come from common observation at actual fire scenes~\cite{Lentini:2006}.  As a result, the observations are typically qualitative in nature.

Previous fire pattern research by the National Institute of Justice (NIJ), the National Institute of Standards and Technology (NIST), and the United States Fire Administration (USFA) has shown that fire patterns provide data useful for the determination of fire origin.  The reports noted the impact of ventilation on the development of the burn patterns~\cite{Shanley:1997,Putorti:1997}. A large number of other factors affect the formation of these patterns: burn time, heat release rate of the fire source, fire exposure, target fuel composition, adjacent fuel(s) and compartmentalization, to name a few. Given the limited number of experiments in the literature and the large number of variables, it has been difficult to fully understand a cause and effect relationship between the fire scenarios and the resulting patterns.  Therefore, fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause~\cite{Sutula}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs). These spread sheets incorporate references to the source documents and experimental work upon which the correlations are based. In addition, NRC tasked NIST with conducting a verification and validation study of the correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study because an insufficient amount of reliable data is available to conduct validation.  Two correlations that could not be validated were for flame height and time to ignition.

In 2003, a list of zone and computational fluid dynamics (CFD) fire models was compiled by Olenick and Carpenter~\cite{Olenick:2003}. A similar survey was conducted in 1992, and Olenick has updated the list in 2014. There are 53 zone models listed, but only three are actively supported or maintained.  The survey also includes 22 field models of which 10 are listed as being actively supported.  All of the correlations and models have been to some extent been validated or checked against experimental data for a limited set of scenarios~\cite{ASTM_E1355}.

Of the active field models, FDS version 6 has one of the most rigorous and publicly accessible verification and validation programs~\cite{McGrattan:2014}. More than 45 experimental data sets are used to quantify the prediction capabilities of the model.  Results are presented in terms of model bias and variation to inform users of whether particular quantities (temperature, heat flux, pressure, etc.) are over or under predicted compared to experimental data.

However for fire models to be demonstrated as reliable and suitable for use in court, studies such as the one conducted here need to be conducted to provide the scientific basis for the use of fire models for re-construction.  The research results may enable fire investigation teams to have a better understanding of the fire scene and enable improved support of their hypothesis on the area of origin based on fire patterns.  This idea was discussed when the Fire Protection Research Foundation convened a Research Advisory Council on Post-fire Analysis in 2002.  Recommendations for research and development in their white paper included, ``advance the capabilities of computational fluid dynamics fire modeling, particularly as applied to fire ignition scenarios and fire pattern development and interpretation''~\cite{RAC:2002}.

Given that fire pattern interpretation and analysis is a critical step in most fire investigations, it is surprising to find the limited amount or research and analysis tools directly related to this practice.  The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA~921, {\em Guide for Fire and Explosion Investigations}~\cite{NFPA_921}.  The first edition of NFPA~921 was issued in 1992.  As provided by the document scope, it ``is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.'' The document’s purpose includes the goal to be ``a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.'' However, NFPA~921 can only incorporate the fire science and research results that have been produced.  Although NFPA~921 is a ``guide'' as opposed to a ``standard,'' it is considered the standard of care for the fire investigation community in countries that recognize and use NFPA standards and guides.

As the result of the review of several arson cases, most notably the cases of Ernest Ray Willis and Cameron Todd Willingham, the practice of fire investigation as a forensic science was called into question.  Beyler conducted a review of both cases for the Texas Forensic Science Commission.  Both fires occurred prior to the release of NFPA~921 in 1992.  The Beyler analysis, issued in August of 2009, concluded that the investigations of both cases ``did not comport with either the modern standard of care expressed by NFPA 921, or the standard of care expressed by fire investigation texts and papers in the period 1980 – 1992.'' The fire investigators had a poor understanding of science and their findings of arson could not be sustained~\cite{Beyler:2009}.

Another landmark document regarding the state of forensic science in the United States was published in 2009.  The 328 page report, entitled {\em Strengthening Forensic Science in the United States: A Path Forward}, by the U.S.~National Academy of Sciences, reviewed 13 different forensic science disciplines including biological evidence, friction ridge patterns, tool mark and firearms identification, and analysis of explosives evidence and fire debris.  The review of the analysis of explosives evidence and fire debris is covered in three pages of the report.  The summary assessment of analysis of explosives evidence is positive, ``the scientific foundations exist to support the analysis of explosions, because such analysis is based primarily on well-established chemistry ''~\cite{Forensic:2009}.  The complete summary assessment for fire debris analysis is included here: ``By contrast, much more research is needed on the natural variability of burn patterns and damage characteristics and how they are affected by the presence of various accelerants.  Despite the paucity of research, some arson investigators continue to make determinations about whether or not a particular fire was set.  However, according to testimony presented to the committee, many of the rules of thumb, that were typically assumed to indicate that an accelerant was used (e.g. ``alligatoring'' of wood, specific char patterns) have been shown not to be true.  Experiments should be designed to put arson investigations on a more solid scientific footing .''
This study is in response to the assessment above, specifically regarding the natural variability of fire patterns.

@@

Well-documented and characterized technologies such as thermocouples and heat flux gauges were used to assess the fire and the thermal exposure of the target material.  Oxygen consumption calorimetry and mass loss measurements were used for the determining the heat release rate of the fires.

The most important measurements for the field investigator are the visual measurements that can be made at the scene or determined from scene photographs. Several methods of imaging were used for these experiments: still photography, digital video, and infrared video.  Computer based imaging techniques, such as those contained in the Image Stream program suite, were examined and adapted to analyze the flame height of the source fires~\cite{Nokes:2011}.

@@

The fires were characterized in terms of heat release rate, mass loss rate, fire plume temperatures and emitted heat flux. Videos were recorded for each experiment.  This ``field'' data from the videos was  used to determine the flame heights. Experiments were conducted with three different fuel types: a gas fueled mass flow controlled burner, a liquid fueled burner, and a solid fuel.

\subsubsection{Instrumented Calibration Target Wall}

A free standing, non-combustible, calcium silicate wall was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.

@@

Comparative experiments were conducted to examine the impact of the method of construction and support of the walls and corner sections on the development of the burn patterns.  Three cases were examined: (1) 12.7~mm thick, regular core, painted gypsum board on the front of a wall frame (open back); (2) 12.7~mm thick, regular core, painted gypsum board on the front and back of a wall frame (interior wall); and {3) 12.7~mm thick, regular core, painted gypsum board on the front and back of a wall frame, with fiberglass batting insulation in the voids of the wood frame (exterior wall).  The main consideration was to identify any differences in the height, width, shape, and area of the fire patterns. The results of these experiments determined the construction of the walls and corners used in subsequent phases of this study.

@@

Several types of bench scale experiments were conducted to generate data. This data could be used to assist in the analysis of the fire patterns and potentially be used as input for the FDS models.
Critical Heat Flux – Conducted a series of experiments on samples of painted gypsum wallboard to determine the repeatability of a ``critical heat flux'' that results in a clear line of demarcation, a ``burn/no burn'' line.  The flooring radiant panel (ASTM~E648) was used to examine this value.

@@

Cone calorimeter (ASTM~1354) experiments were conducted to estimate the ignition temperature, the time to ignition and the heat release rate from samples of painted gypsum wallboard under thermal flux conditions ranging from 25~kW/m$^2$ to 75~kW/m$^2$.
